---
title: "Assignment 3"
author: "Robert Isdell"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(automap)
library(dismo)
library(gstat) # Spatial and Spatio-Temporal Geostatistical Modelling, Prediction and Simulation
library(maptools) # Tools for Handling Spatial Objects
library(Matrix)
library(raster) # Geographic Data Analysis and Modeling
library(rgdal) # Bindings for the 'Geospatial' Data Abstraction Library
library(rstan)
library(sf) # Simple Features for R
library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(tmaptools) # Thematic Map Tools
```

### Load the data

```{r}
dat <- read_sf(here::here("Data/GIS/FLMarshPearlsUpdated.shp"))
```

### Quick plot

```{r}
plot(dat["Pearls"], pch = 16, cex = 2)
```

### Project to useful coordinates

```{r}
nad83 <- st_crs("EPSG:3968")
d2 <- st_transform(dat, crs = nad83)
```

### Convert to SpatialPointsDataframe

Because gstat is obnoxious.

```{r}
d3 <- as(d2, "Spatial")
```


## Create a Grid

```{r}
siteGrid <- raster(
  d2,
  res = 0.25,
  vals = 1
)
sitePoints <- as.data.frame(rasterToPoints(siteGrid))
sitePoints <- SpatialPixelsDataFrame(
  points = sitePoints[c("x", "y")],
  data = sitePoints
)
gridded(sitePoints) <- TRUE
fullgrid(sitePoints) <- TRUE
proj4string(sitePoints) <- proj4string(siteGrid)
```

## Inverse distance weighting

```{r, message=FALSE}
pearlIDW <- gstat::idw(
  formula = Pearls ~ 1,
  locations = d3,
  newdata = sitePoints,
  idp = 0.5
)
spplot(pearlIDW, "var1.pred", main = "Power = 0.5")
pearlIDW1 <- gstat::idw(
  formula = Pearls ~ 1,
  locations = d3,
  newdata = sitePoints,
  idp = 1
)
spplot(pearlIDW1, "var1.pred", main = "Power = 1")
pearlIDW2 <- gstat::idw(
  formula = Pearls ~ 1,
  locations = d3,
  newdata = sitePoints,
  idp = 2
)
spplot(pearlIDW2, "var1.pred", main = "Power = 2")
pearlIDW3 <- gstat::idw(
  formula = Pearls ~ 1,
  locations = d3,
  newdata = sitePoints,
  idp = 3
)
spplot(pearlIDW3, "var1.pred", main = "Power = 3")
```

## Kriging

### Build a variogram

```{r}
PearlVar <- variogram(Pearls~1, d3, alpha = 45, width = 1)
summary(PearlVar)
plot(
  PearlVar,
  pch = 20,
  cex = 1.5,
  col = "black",
  ylab = expression("Semivariance ("*gamma*")"),
  xlab = "Distance (m)",
  main = "Pearl Counts"
)
```

### Fit a model

```{r}
pow.model <- vgm(
  model = "Pow",
  range = 1
)
pow.fit <- fit.variogram(
  object = PearlVar,
  model = pow.model
)
plot(
  PearlVar,
  pch = 20,
  cex = 1.5,
  col = "black",
  ylab = expression("Semivariance ("*gamma*")"),
  xlab = "Distance (m)",
  main = "Pearl Counts",
  model = pow.fit
)

Exp.model <- vgm(
  model = "Exp",
  range = 1
)
Exp.fit <- fit.variogram(
  object = PearlVar,
  model = Exp.model
)
plot(
  PearlVar,
  pch = 20,
  cex = 1.5,
  col = "black",
  ylab = expression("Semivariance ("*gamma*")"),
  xlab = "Distance (m)",
  main = "Pearl Counts",
  model = Exp.fit
)
```

Select the power model.

### Interpolate

```{r, message=FALSE}
PearlPow <- krige(
  formula = Pearls~1,
  d3,
  sitePoints,
  model = pow.fit
)
spplot(
  PearlPow,
  "var1.pred"
)
```

### Automap version

```{r, warning=FALSE, message=FALSE}
PearlVarAuto <- autofitVariogram(Pearls~1, d3)
plot(PearlVarAuto)
PearlKrig <- autoKrige(Pearls~1, d3, sitePoints)
plot(PearlKrig)
plot(PearlKrig$krige_output)
```

# Stan model

## Convert the points to polygons

```{r}
PearlPolys <- dismo::voronoi(d3)
```

## Prep the data

### Load the functions

Functions obtained from [Connor Donegan](https://github.com/ConnorDonegan/Stan-IAR/blob/main/icar-functions.R).

```{r}
#' connect_regions
#' 
#' Given an nb object and the names of two areal regions, update the nb
#' object so that the two regions are connected.

#' The nb object is a list of n integer vectors.  It also has attribute
#' region.id which is a character vector with n unique values (like the
#' row.names of a data.frame object); n is the number of spatial entities.
#' Component i of this list contains the integer identifiers its neighbours
#' as a sorted vector with no duplication and values in 1:n;  if i has no
#' neighbours, the component is a vector of length 1 with value 0L.
#' see:  https://cran.r-project.org/web/packages/spdep/vignettes/nb_igraph.html
#'
#' param nb: nb object over areal regions
#' param name1:  name of region 1.
#' param name2:  name of region 1.
#' returns: updated nb object
#' 
#' @author Mitzi Morris
#' 
connect_regions <- function(nb, name1, name2) {
  if (name1 == name2) {
    cat("Cannot connect region to itself: ", name1)
    return(nb)
  }
  id1 <- which(attr(nb, "region.id") == name1)
  if (length(id1) == 0) {
    cat("Unknown region: ", name1)
    return(nb)
  }
  id2 <- which(attr(nb, "region.id") == name2)
  if (length(id2) == 0) {
    print("Unknown region: ", name2)
    return(nb);
  }
  if (nb[[id1]][1] == 0)  # singleton
    nb[[id1]] <- c(as.integer(id2))
  else
    nb[[id1]] <- unique(sort(c(nb[[id1]], as.integer(id2))))
  
  if (nb[[id2]][1] == 0)  # singleton
    nb[[id2]] <- c(as.integer(id1))
  else
    nb[[id2]] <- unique(sort(c(nb[[id2]], as.integer(id1))))
  nb
}


#' convert connectivity matrix to unique pairs of connected nodes (graph structure)
#' 
#' @param w a connectivity matrix
#' 
#' @return a data.frame with three columns: node1 and node2 (the indices of connected nodes) and their weight (the element w[i,j]).
#'   Only unique pairs of connected nodes are included---that is, each pair `[i,j]` is listed once, 
#'    with all i < j. This means that if `[i, j]` is included, then `[j, i]` is not also listed.
#'    
#' @author Connor Donegan
#' 
edges <- function (w) {
  lw <- apply(w, 1, function(r) {
    which(r != 0)
  })
  all.edges <- lapply(1:length(lw), function(i) {
    nbs <- lw[[i]]
    if (length(nbs)) 
      data.frame(node1 = i, node2 = nbs, weight = w[i, nbs])
  })
  all.edges <- do.call("rbind", all.edges)
  edges <- all.edges[which(all.edges$node1 < all.edges$node2), ]
  return(edges)
}


#' compute scaling factor for adjacency matrix
#' accounts for differences in spatial connectivity 
#' 
#' @param C connectivity matrix
#' 
#' Requires the following packages: 
#' 
#' library(Matrix)
#' library(INLA);
#' library(spdep)
#' library(igraph)
#' 
#' @author Mitzi Morris
#' 
scale_c <- function(C) {
  #' compute geometric mean of a vector
  geometric_mean <- function(x) exp(mean(log(x))) 
  
  N = dim(C)[1]
  
  # Create ICAR precision matrix  (diag - C): this is singular
  # function Diagonal creates a square matrix with given diagonal
  Q =  Diagonal(N, rowSums(C)) - C
  
  # Add a small jitter to the diagonal for numerical stability (optional but recommended)
  Q_pert = Q + Diagonal(N) * max(diag(Q)) * sqrt(.Machine$double.eps)
  
  # Function inla.qinv provides efficient way to calculate the elements of the
  # the inverse corresponding to the non-zero elements of Q
  Q_inv = inla.qinv(Q_pert, constr=list(A = matrix(1,1,N),e=0))
  
  # Compute the geometric mean of the variances, which are on the diagonal of Q.inv
  scaling_factor <- geometric_mean(Matrix::diag(Q_inv)) 
  return(scaling_factor) 
}

#' prepare Stan data for ICAR model given a connectivity matrix
#' 
#' @param C a connectivity matrix
#' @param scale_factor optional vector of scale factors for each connected portion of the graph structure. 
#'   Generally, you will ignore this and update the scale factor manually.
#'   
#' @return a list with all that is needed for the Stan ICAR prior. If you do not provide inv_sqrt_scale_factor, 
#'   it will be set to a vector of 1s.
#'   
#' @author Connor Donegan
#' 
prep_icar_data <- function (C, inv_sqrt_scale_factor = NULL) {
  n <- nrow(C)
  E <- edges(C)
  G <- list(np = nrow(C), from = E$node1, to = E$node2, nedges = nrow(E))
  class(G) <- "Graph"
  nb2 <- spdep::n.comp.nb(spdep::graph2nb(G))
  k = nb2$nc
  if (inherits(inv_sqrt_scale_factor, "NULL")) inv_sqrt_scale_factor <- array(rep(1, k), dim = k)
  group_idx = NULL
  for (j in 1:k) group_idx <- c(group_idx, which(nb2$comp.id == j))
  group_size <- NULL
  for (j in 1:k) group_size <- c(group_size, sum(nb2$comp.id == j))
  # intercept per connected component of size > 1, if multiple.
  m <- sum(group_size > 1) - 1
  if (m) {
    GS <- group_size
    ID <- nb2$comp.id
    change.to.one <- which(GS == 1)
    ID[which(ID == change.to.one)] <- 1
    A = model.matrix(~ factor(ID))
    A <- as.matrix(A[,-1])
  } else {
    A <- model.matrix(~ 0, data.frame(C))
  }
  l <- list(k = k, 
            group_size = array(group_size, dim = k), 
            n_edges = nrow(E), 
            node1 = E$node1, 
            node2 = E$node2, 
            group_idx = array(group_idx, dim = n), 
            m = m,
            A = A,
            inv_sqrt_scale_factor = inv_sqrt_scale_factor, 
            comp_id = nb2$comp.id)
  return(l)
}

#' Moran Coefficient
#' 
#' @param x vector of numeric values 
#' @param w connectivity matrix 
#' 
#' @author Connor Donegan
mc <- function (x, w, digits = 3, warn = TRUE) {
  if (missing(x) | missing(w)) 
    stop("Must provide data x (length n vector) and n x n spatial weights matrix (w).")
  if (any(rowSums(w) == 0)) {
    zero.idx <- which(rowSums(w) == 0)
    if (warn) 
      message(length(zero.idx), " observations with no neighbors found. They will be dropped from the data.")
    x <- x[-zero.idx]
    w <- w[-zero.idx, -zero.idx]
  }
  xbar <- mean(x)
  z <- x - xbar
  ztilde <- as.numeric(w %*% z)
  A <- sum(rowSums(w))
  n <- length(x)
  mc <- as.numeric(n/A * (z %*% ztilde)/(z %*% z))
  return(round(mc, digits = digits))
}
```

### Process the data

```{r}
C <- spdep::nb2mat(spdep::poly2nb(PearlPolys, queen = TRUE),
                   style = "B", zero.policy = TRUE)
icar.data <- prep_icar_data(C)
```

## Run the model

```{r}
Xmat <- model.matrix(lm(Pearls~Mussels, data = d3))

standat <- list(
  n = 20,
  m = 2,
  Xmat = Xmat,
  count = round(d3$Pearls),
  N_edges = icar.data$n_edges,
  node1 = icar.data$node1,
  node2 = icar.data$node2
)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

mod <- stan_model("Scripts/Assignment3.stan")
mod <- readRDS(here::here("Scripts/Assignment3.rds"))

fit <- sampling(mod,
                data = standat,
                iter = 50000,  ## `iter` _includes_ `warmup` count
                warmup = 40000,
                chains = 5,  ## default is 4
                thin = 2, ## default is 1
                control = list(adapt_delta = .9,
                               max_treedepth = 15)
                ## https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup
)
saveRDS(fit, "Assigment3Posteriors.rds")
traceplot(fit, pars=c("beta", "sigma", "phi"))
plot(fit, pars=c("beta", "sigma"))
```


## NIMBLE

```{r}
W.nb <- poly2nb(PearlPolys, row.names =  rownames(PearlPolys@data))
## Determine neighborhood/adjacency information needed for neighborhood-based CAR model
nbInfo <- nb2WB(W.nb)
```

```{r}
nregions <- nrow(PearlPolys)

PearlConstants <- list(
  n = nregions, 
  L = length(nbInfo$adj), 
  adj = nbInfo$adj, 
  weights = nbInfo$weights, 
  num = nbInfo$num,
  mussels = as.vector(scale(PearlPolys$Mussels))
)

NimbleData <- list(
  count = round(PearlPolys$Pearls)
)

NimbleInits <- list(beta0 = 0, beta1 = 0, sigma = 1, s = rnorm(nregions))

m1 <- nimbleCode({
  # Priors
  beta0 ~ dnorm(0, 100)
  beta1 ~ dnorm(0, 5)
  sigma ~ dexp(1)
  tau <- 1/sigma^2
  rate ~ dexp(1)
  # latent process
  s[1:n] ~ dcar_normal(adj[1:L], weights[1:L], num[1:n], tau, zero_mean = 0)
  # likelihood
  for(j in 1:n) {
    count[j] ~ dpois(lambda[j])
    log(lambda[j]) <- beta0 + beta1*mussels[j] + s[j]
  }
})

PearlModel <- nimbleModel(
  m1, 
  constants = PearlConstants, 
  data = NimbleData, 
  inits = NimbleInits
)

cModel <- compileNimble(PearlModel)

# Check the samplers
conf <- configureMCMC(PearlModel, monitors = c('beta0', 'beta1', 'sigma', 's'))
conf$printSamplers()

MCMC <- buildMCMC(conf)
cMCMC <- compileNimble(MCMC, project = cModel)

samples <- runMCMC(
  cMCMC, 
  niter = 100000, 
  nburnin = 90000,
  thin = 2)
```

