---
title: "tmap Example"
author: "Robert Isdell"
date: "`r Sys.Date()`"
output: html_document
---

# Example of plotting spatial data using the `tmap` package.

## Load the necessary packages.
```{r setup, message=FALSE, warning=FALSE}
library(rgdal) # Bindings for the 'Geospatial' Data Abstraction Library
library(sf) # Simple Features for R
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(tmaptools) # Thematic Map Tools
```

## Load the data
```{r}
dat <- sf::read_sf(here::here("Data/GIS/PlantDensity.shp"))
```

## Add a basemap
```{r}
bmap <- tmaptools::read_osm(dat, 
                            ext = 1.1) # add 10% to bounding box
```

This function pulls the Open Street Map basemap tiles, composites them based on the extent of your input layer, and then saves the input as a raster to add under your features.

## Create the map

### Static

```{r}
tmap_mode(mode = "plot")
tmap::tm_shape(bmap) + #in tmap, everything starts as a "shape"
  tm_rgb() + # this is for multiband display, default options
  tm_shape(dat) + # Add the next "shape"
  tm_bubbles(size = "Stems", # bubbles are useful for displaying sizes
             shape = 21, # pick from the standard pch symbols
             col = "green",
             border.col = "black") + # add a border color
  tm_scale_bar(width = 0.2, # add a scale bar; width in relative units
               text.size = 2,
               lwd = 1,
               position = c("RIGHT", "BOTTOM")) + # Caps for closer to margin
  tm_compass(type = "arrow", # Add an arrow compass
             size = 4,
             position = c(0.85, 0.15)) + # relative position
  tm_layout(legend.frame = TRUE, # Add a frame and background
            legend.bg.color = "white",
            legend.title.size = 2,
            legend.text.size = 1)

```

### Interactive

The parts that are commented out aren't valid for viewing mode
```{r}
tmap_mode(mode = "view")
# tmap::tm_shape(bmap) + #in tmap, everything starts as a "shape"
#   tm_rgb() + # this is for multiband display, default options
tm_basemap("Stamen.Watercolor") + 
tm_shape(dat) +
  tm_dots(col = "Stems", 
          size = 0.05,
          shape = 16,
          border.col = "black",
          palette = "inferno") +
  tm_scale_bar(# width = 0.2,
               text.size = 2,
               lwd = 1,
               position = c("left", "bottom")) +
#   tm_compass(type = "arrow", # not supported in view mode
#              size = 3) +
  tm_layout(legend.frame = TRUE,
            legend.bg.color = "white",
            legend.title.size = 2,
            legend.text.size = 1)

```